---
title: "Running Background Jobs with franzpak"
format: html
---

# Introduction

**ŌÜĀ’ĖÅ Warning:** These functions are still evolving and can be considered
experimental.

This vignette shows how to use the background-job functions (`bgjm_*`) to run
long-running statistical models in background R processes, with practical
examples using `lavaan`, `tidyLPA`, and `MplusAutomation`.

Under the hood the jobs run on the [mirai](https://mirai.r-lib.org) framework.
On top of mirai, `franzpak` adds two conveniences:

- a **named registry**, so you start/list/collect jobs by a friendly name
  instead of holding objects yourself, and
- **per-job filesystem management**, so each job gets its own directory, its
  stdout/stderr go to log files, and any files the run writes (e.g. Mplus
  `.inp`/`.dat`/`.out`) are collected into an `artifacts/` folder.

This is useful for:

- freeing your interactive session while compute-heavy models run (bootstrapped
  SEM, mixture models, Mplus runs),
- isolating model execution from your main session, and
- keeping logs and output files organized per job.

If you need the full power of async/parallel R ŌĆö element-wise parallel maps,
progress bars, timeouts, Shiny/promises, remote/HPC ŌĆö use mirai directly. The
`bgjm_*` layer intentionally exposes only the registry + filesystem conveniences.

# Getting started

The job system requires `mirai` (a hard dependency). Optional packages enable
the model launchers:

- `lavaan` ŌĆö structural equation modeling
- `tidyLPA` (>= 1.0.0) + `mclust` ŌĆö latent profile analysis
- `MplusAutomation` (+ an Mplus installation) ŌĆö Mplus models

```{r}
library(franzpak)
library(lavaan)
```

## The daemon pool

Jobs run on a dedicated pool of background workers (mirai daemons). The pool
starts automatically on your first `bgjm_start_*()` call, but you can size it
explicitly ŌĆö and shut it down ŌĆö with `bgjm_daemons()`:

```{r}
#| eval: false
bgjm_daemons(4)   # use 4 background workers
# ... run jobs ...
bgjm_daemons(0)   # shut the pool down
```

The pool also bounds how many jobs run at once: with 4 workers, a fifth
submitted job simply queues until a worker frees up. You can set a session
default with `options(franzpak.bgjm_daemons = 4)`.

## A complete workflow

```{r}
#| eval: false
# Define a CFA model
model_syntax <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
data(HolzingerSwineford1939)

# Launch in the background ŌĆö returns immediately with a job id
job_id <- bgjm_start_lavaan(
  model_syntax = model_syntax,
  data = HolzingerSwineford1939,
  lavaan_fun = "cfa",
  job_name = "holzinger-cfa"
)
```

`bgjm_start_lavaan()` returns at once, **without waiting** for the model to
finish, so your session stays free. The next section covers how to get the
result back.

# Retrieving results

A job hands you back only a *job id* ŌĆö a short string. The fitted model lives in
the background process until you ask for it. There are three ways to interact
with a job, depending on whether you want to wait:

**1. `bgjm_result(job_id)` ŌĆö get the model, waiting if needed.** This is the
usual way to retrieve output. It returns the fitted object directly:

```{r}
#| eval: false
fit <- bgjm_result(job_id)
lavaan::fitMeasures(fit, c("cfi", "rmsea"))

bgjm_remove(job_id, delete_dir = TRUE)   # clean up when done
```

**2. `bgjm_collect(job_id)` ŌĆö get the model *plus* its files and logs.** Use
this when you also need the artifacts directory or log paths. It returns a list:

```{r}
#| eval: false
collected <- bgjm_collect(job_id)
fit       <- collected$result            # the fitted model
collected$artifacts_dir                  # files written during the run
collected$stdout                         # path to captured stdout log
```

The `collected` list contains `result`, `job_dir`, `artifacts_dir`,
`artifact_files`, `stdout`, and `stderr`. (`bgjm_result()` is simply a shortcut
for `bgjm_collect(job_id)$result`.)

**3. `bgjm_status()` / `bgjm_unresolved()` ŌĆö check without waiting.** Use these
when you do *not* want to block ŌĆö e.g. to poll, or to decide whether a result is
ready:

```{r}
#| eval: false
bgjm_status(job_id)        # "running", "finished", or "error"
bgjm_unresolved(job_id)    # TRUE while still running, FALSE once resolved
```

## Blocking is intentional

Both `bgjm_result()` and `bgjm_collect()` **block**: if the job is still
running, the call waits for it to finish rather than returning early or raising
an error. This mirrors mirai's `m[]`. It is what you want when collecting:

- in a **script**, you want the line to wait until the model is ready;
- in a **Quarto / R Markdown report**, blocking is essential ŌĆö otherwise the
  document would render straight past your models before they finish, producing
  empty or missing output.

If you would rather *not* wait, check `bgjm_unresolved(job_id)` first and only
collect once it returns `FALSE`. If a job *failed*, `bgjm_result()` /
`bgjm_collect()` raise an informative error pointing at the job's logs (see
[Handling failures](#handling-failures)).

Convenient one-liner: pass `auto_remove = TRUE` to collect the result and drop
the job (and its working directory) from the registry in one step:

```{r}
#| eval: false
fit <- bgjm_result(job_id, auto_remove = TRUE)
```

# Examples

## SEM mediation with bootstrapped standard errors

Bootstrap procedures are ideal background jobs given their cost:

```{r}
#| eval: false
mediation_model <- "
  M ~ a * X
  Y ~ b * M + c * X
  indirect := a * b
  total := c + (a * b)
"

set.seed(123)
n <- 200
X <- rnorm(n); M <- 0.5 * X + rnorm(n); Y <- 0.3 * M + 0.2 * X + rnorm(n)
mediation_data <- data.frame(X, M, Y)

job_id <- bgjm_start_lavaan(
  model_syntax = mediation_model,
  data = mediation_data,
  lavaan_fun = "sem",
  job_name = "mediation",
  se = "bootstrap",
  bootstrap = 1000,
  seed = 42                # reproducible bootstrap
)

# ... keep working ...

fit <- bgjm_result(job_id, auto_remove = TRUE)
lavaan::parameterEstimates(fit, boot.ci.type = "bca.simple")
```

## Latent profile analysis with tidyLPA

```{r}
#| eval: false
library(tidyLPA)

job_id <- bgjm_start_tidylpa(
  data = iris[, 1:4],
  n_profiles = 3,
  models = 1,
  package = "mclust",
  job_name = "iris-lpa"
)

lpa <- bgjm_result(job_id, auto_remove = TRUE)
lpa
```

## Several models at once

Because jobs are non-blocking, launch competing models together and collect them
afterwards. The daemon pool caps how many actually run in parallel. With
`purrr`, starting and collecting are each one line:

```{r}
#| eval: false
library(purrr)

models <- list(
  "1-factor" = "g =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9",
  "2-factor" = "verbal =~ x1 + x2 + x3 + x4 + x5 + x6\n speed =~ x7 + x8 + x9",
  "3-factor" = "visual =~ x1 + x2 + x3\n textual =~ x4 + x5 + x6\n speed =~ x7 + x8 + x9"
)

# Start every model in the background; imap passes each syntax (.x) and its
# name (.y). Returns a named vector of job ids.
jobs <- imap_chr(models, \(syntax, name)
  bgjm_start_lavaan(syntax, HolzingerSwineford1939, "cfa", job_name = name))

# Collect them ŌĆö each call waits until that job is ready.
fits <- map(jobs, bgjm_result)

tibble::tibble(
  model = names(fits),
  cfi   = map_dbl(fits, lavaan::fitMeasures, "cfi"),
  rmsea = map_dbl(fits, lavaan::fitMeasures, "rmsea"),
  aic   = map_dbl(fits, lavaan::fitMeasures, "aic")
)

walk(jobs, bgjm_remove, delete_dir = TRUE)
```

# Mplus jobs and file artifacts

Some tools ŌĆö Mplus above all ŌĆö produce *files* (`.inp`, `.dat`, `.out`). The job
system captures these for you. When a job runs, its directory looks like:

```
job_directory/
Ōö£ŌöĆŌöĆ tmp/              # scratch (TMPDIR) for the job
Ōö£ŌöĆŌöĆ artifacts/        # files written during the run land here
Ōö£ŌöĆŌöĆ stdout.log        # captured standard output
ŌööŌöĆŌöĆ stderr.log        # captured messages/warnings
```

## Running an Mplus input: `bgjm_start_mplus()`

`bgjm_start_mplus()` runs an existing Mplus input via
`MplusAutomation::runModels()`. Point it at a `.inp` file or a directory holding
the `.inp` (and its `.dat`); the inputs are staged into the job directory, Mplus
runs there, and the `.out` plus an `mplus_run.log` are captured as artifacts. The
collected `$result` is the parsed `MplusAutomation::readModels()` object.

```{r}
#| eval: false
job_id <- bgjm_start_mplus(
  target = "path/to/model_dir",   # holds model.inp (+ model.dat)
  job_name = "my-mplus-run"
)

out <- bgjm_collect(job_id)

list.files(out$artifacts_dir)     # e.g. "model.out", "mplus_run.log"
out$result                        # parsed model (readModels)

# Mplus streams its own console log (showOutput = TRUE, quiet = FALSE);
# on macOS/Linux it is captured to the job's stdout log:
head(bgjm_read_stdout(job_id))
```

To keep the output files permanently, pass a persistent `output_base` and remove
the job from the registry *without* deleting its directory:

```{r}
#| eval: false
job_id <- bgjm_start_mplus(
  target = "path/to/model_dir",
  output_base = file.path(getwd(), "mplus_results")
)
out <- bgjm_collect(job_id)
bgjm_remove(job_id, delete_dir = FALSE)   # keep files on disk
```

By default `output_base` is R's `tempdir()`, which may be cleared when R exits ŌĆö
use a persistent `output_base` for anything you want to keep.

## LPA via the Mplus backend

`bgjm_start_tidylpa(package = "Mplus")` runs LPA through Mplus and captures the
same artifacts. Mplus requires alphanumeric/underscore variable names (no dots):

```{r}
#| eval: false
iris_mplus <- iris[, 1:4]
names(iris_mplus) <- c("Sepal_Length", "Sepal_Width",
                       "Petal_Length", "Petal_Width")

job_id <- bgjm_start_tidylpa(
  data = iris_mplus, n_profiles = 3, models = 1,
  package = "Mplus", keepfiles = TRUE,
  job_name = "iris-lpa-mplus"
)
result <- bgjm_collect(job_id)
list.files(result$artifacts_dir)
```

# Parallelism and CPU budgeting

Two things can run in parallel at the same time, and it is easy to accidentally
ask for far more CPU than you have:

1. **The pool** runs several *jobs* at once ŌĆö `bgjm_daemons(W)` gives `W` worker
   processes (the *outer* layer).
2. **A single job** can itself use several cores ŌĆö lavaan can parallelise a
   bootstrap, and Mplus can use multiple processors per run (the *inner* layer).

If `W` workers each use `C` cores, peak demand is roughly **`W ├Ś C` cores**. Push
that above your physical core count and everything slows down through
contention. The rule of thumb:

> Keep `workers ├Ś cores-per-job Ōēż physical cores`, leaving one or two free for
> the OS and your interactive session.

```{r}
#| eval: false
parallelly::availableCores()   # or parallel::detectCores()
```

There are two sensible strategies; pick based on whether you have *many* fits or
*a few big* ones.

## Strategy A ŌĆö many single-threaded jobs (maximise throughput)

When you have lots of independent fits (model comparison, many bootstrap
replications split across jobs), give all cores to the *pool* and keep each job
single-threaded. With 14 cores:

```{r}
#| eval: false
bgjm_daemons(12)               # 12 workers, each using 1 core
# launch many jobs; the pool runs 12 at a time
```

## Strategy B ŌĆö a few multi-threaded jobs (minimise latency per fit)

When you have one or a few expensive models and want each to finish fast, keep
the pool small and let each job parallelise internally so that
`W ├Ś C` still fits your budget (e.g. `3 ├Ś 4 = 12`):

```{r}
#| eval: false
bgjm_daemons(3)                # 3 workers ...
```

### lavaan bootstrap inside a job

lavaan's bootstrap accepts `parallel`, `ncpus`, and `iseed`, and these pass
straight through `bgjm_start_lavaan()`'s `...`:

```{r}
#| eval: false
job <- bgjm_start_lavaan(
  mediation_model, mediation_data, "sem",
  se = "bootstrap", bootstrap = 1000,
  parallel = "snow",           # see note below
  ncpus = 4,                   # 4 cores for this one job
  iseed = 123,                 # reproducible *parallel* bootstrap
  job_name = "med-boot"
)
```

Two notes:

- **Prefer `parallel = "snow"` over `"multicore"` inside background jobs.** Jobs
  already run in their own worker process; `"snow"` spawns fresh child processes,
  whereas `"multicore"` *forks* the worker ŌĆö and forking from a process that has
  background threads (as the workers do) is fragile. `"snow"` is the safe choice
  here.
- For reproducibility of a *parallel* bootstrap, set lavaan's `iseed`. The
  `seed` argument of `bgjm_start_*()` seeds the job's main process, which is not
  enough once the bootstrap itself runs on several cores.

### Mplus processors inside a job

Mplus parallelises a single run via the `PROCESSORS` option in the input file's
`ANALYSIS:` block ŌĆö set it in the `.inp` you pass to `bgjm_start_mplus()`:

```
ANALYSIS:
  PROCESSORS = 4;
```

Then keep the pool small (`bgjm_daemons(3)` with `PROCESSORS = 4` Ōēł 12 cores).
Because Mplus is an external program, there is no forking concern ŌĆö it manages
its own threads.

## The hidden third layer: multithreaded BLAS

Some builds of R use a multithreaded BLAS (OpenBLAS, Apple Accelerate, MKL). If
each worker's BLAS also grabs several threads, you can oversubscribe without
realising it. When running many workers, it is usually best to pin each worker's
BLAS to a single thread:

```{r}
#| eval: false
bgjm_daemons(12)
# pin BLAS to 1 thread on every franzpak worker
mirai::everywhere(
  try(RhpcBLASctl::blas_set_num_threads(1L), silent = TRUE),
  .compute = "franzpak"
)
```

This keeps your `W ├Ś C` budget honest and is rarely a downside for the kinds of
models run here (lavaan ML, mixture models), which are not BLAS-bound.

# Monitoring, logs, and control

```{r}
#| eval: false
bgjm_list()                       # registry snapshot (id, name, status, paths)

bgjm_status(job_id)               # non-blocking status

cat(bgjm_read_stdout(job_id), sep = "\n")   # captured stdout (incl. Mplus log)
cat(bgjm_read_stderr(job_id), sep = "\n")   # captured messages/warnings

bgjm_kill(job_id)                 # cancel a running job (mirai::stop_mirai)
bgjm_remove(job_id, delete_dir = TRUE)      # drop from registry (+ files)
```

## Job naming

A supplied `job_name` is used verbatim when unique, otherwise `-1`, `-2`, ... is
appended. With no name, a sensible default is generated (`"lavaan-cfa"`,
`"tidylpa-n3-mclust"`, `"mplus"`):

```{r}
#| eval: false
bgjm_start_lavaan(model_syntax, data, "cfa", job_name = "my-model")  # "my-model"
bgjm_start_lavaan(model_syntax, data, "cfa", job_name = "my-model")  # "my-model-1"
```

## Handling failures

A job that errors reports `status == "error"`, and `bgjm_collect()` raises an
error pointing at the logs:

```{r}
#| eval: false
result <- tryCatch(
  bgjm_collect(job_id),
  error = function(e) {
    message("Job failed: ", conditionMessage(e))
    cat(bgjm_read_stderr(job_id), sep = "\n")
    NULL
  }
)
bgjm_remove(job_id, delete_dir = TRUE)
```

# Best practices

- **Clean up in functions** with `on.exit()`:

  ```{r}
  #| eval: false
  analyze <- function(model, data) {
    job <- bgjm_start_lavaan(model, data, "cfa")
    on.exit(bgjm_remove(job, delete_dir = TRUE), add = TRUE)
    bgjm_result(job)
  }
  ```

- **Use persistent `output_base`** for Mplus / anything whose files you want to
  keep; keep `tempdir()` for quick exploratory fits you collect immediately.
- **Use descriptive `job_name`s** so `bgjm_list()` is readable.
- **Size the pool** with `bgjm_daemons(n)` to match your cores ŌĆö and mind
  *nested* parallelism (lavaan bootstrap, Mplus `PROCESSORS`) so that
  workers ├Ś cores-per-job stays within your CPU budget. See
  [Parallelism and CPU budgeting](#parallelism-and-cpu-budgeting).

# Notes and limits

- The registry lives in memory and is lost when R exits (mirai daemons are also
  shut down on package unload). "Background" means *another process this
  session*, not a job that survives a restart.
- `bgjm_collect()` blocks; use `bgjm_status()` / `bgjm_list()` for non-blocking
  checks.
- The Mplus console log is captured to stdout only where `runModels()` routes it
  through R (macOS/Linux). The authoritative record is always the `.out`
  artifact.
- For parallel *maps* with progress bars / early-stop, or timeouts, Shiny, and
  remote execution, prefer `mirai` directly.

# Session info

```{r}
#| eval: true
sessionInfo()
```
