---
title: "Running Background Jobs with franzpak"
format: html
---

# Introduction

**⚠️ Warning:** These functions are still evolving and can be considered
experimental.

This vignette shows how to use the background-job functions (`bgjm_*`) to run
long-running statistical models in background R processes, with practical
examples using `lavaan`, `tidyLPA`, and `MplusAutomation`.

Under the hood the jobs run on the [mirai](https://mirai.r-lib.org) framework.
On top of mirai, `franzpak` adds two conveniences:

- a **named registry**, so you start/list/collect jobs by a friendly name
  instead of holding objects yourself, and
- **per-job filesystem management**, so each job gets its own directory, its
  stdout/stderr go to log files, and any files the run writes (e.g. Mplus
  `.inp`/`.dat`/`.out`) are collected into an `artifacts/` folder.

This is useful for:

- freeing your interactive session while compute-heavy models run (bootstrapped
  SEM, mixture models, Mplus runs),
- isolating model execution from your main session, and
- keeping logs and output files organized per job.

If you need the full power of async/parallel R — element-wise parallel maps,
progress bars, timeouts, Shiny/promises, remote/HPC — use mirai directly. The
`bgjm_*` layer intentionally exposes only the registry + filesystem conveniences.

# Getting started

The job system requires `mirai` (a hard dependency). Optional packages enable
the model launchers:

- `lavaan` — structural equation modeling
- `tidyLPA` (>= 1.0.0) + `mclust` — latent profile analysis
- `MplusAutomation` (+ an Mplus installation) — Mplus models

```{r}
library(franzpak)
library(lavaan)
```

## The daemon pool

Jobs run on a dedicated pool of background workers (mirai daemons). The pool
starts automatically on your first `bgjm_start_*()` call, but you can size it
explicitly — and shut it down — with `bgjm_daemons()`:

```{r}
#| eval: false
bgjm_daemons(4)   # use 4 background workers
# ... run jobs ...
bgjm_daemons(0)   # shut the pool down
```

The pool also bounds how many jobs run at once: with 4 workers, a fifth
submitted job simply queues until a worker frees up. You can set a session
default with `options(franzpak.bgjm_daemons = 4)`.

## A complete workflow

```{r}
#| eval: false
# Define a CFA model
model_syntax <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
data(HolzingerSwineford1939)

# Launch in the background — returns immediately with a job id
job_id <- bgjm_start_lavaan(
  model_syntax = model_syntax,
  data = HolzingerSwineford1939,
  lavaan_fun = "cfa",
  job_name = "holzinger-cfa"
)
```

`bgjm_start_lavaan()` returns at once, **without waiting** for the model to
finish, so your session stays free. The next section covers how to get the
result back.

# Retrieving results

A job hands you back only a *job id* — a short string. The fitted model lives in
the background process until you ask for it. There are three ways to interact
with a job, depending on whether you want to wait:

**1. `bgjm_result(job_id)` — get the model, waiting if needed.** This is the
usual way to retrieve output. It returns the fitted object directly:

```{r}
#| eval: false
fit <- bgjm_result(job_id)
lavaan::fitMeasures(fit, c("cfi", "rmsea"))

bgjm_remove(job_id, delete_dir = TRUE)   # clean up when done
```

**2. `bgjm_collect(job_id)` — get the model *plus* its files and logs.** Use
this when you also need the artifacts directory or log paths. It returns a list:

```{r}
#| eval: false
collected <- bgjm_collect(job_id)
fit       <- collected$result            # the fitted model
collected$artifacts_dir                  # files written during the run
collected$stdout                         # path to captured stdout log
```

The `collected` list contains `result`, `job_dir`, `artifacts_dir`,
`artifact_files`, `stdout`, and `stderr`. (`bgjm_result()` is simply a shortcut
for `bgjm_collect(job_id)$result`.)

**3. `bgjm_status()` / `bgjm_unresolved()` — check without waiting.** Use these
when you do *not* want to block — e.g. to poll, or to decide whether a result is
ready:

```{r}
#| eval: false
bgjm_status(job_id)        # "running", "finished", or "error"
bgjm_unresolved(job_id)    # TRUE while still running, FALSE once resolved
```

## Blocking is intentional

Both `bgjm_result()` and `bgjm_collect()` **block**: if the job is still
running, the call waits for it to finish rather than returning early or raising
an error. This mirrors mirai's `m[]`. It is what you want when collecting:

- in a **script**, you want the line to wait until the model is ready;
- in a **Quarto / R Markdown report**, blocking is essential — otherwise the
  document would render straight past your models before they finish, producing
  empty or missing output.

If you would rather *not* wait, check `bgjm_unresolved(job_id)` first and only
collect once it returns `FALSE`. If a job *failed*, `bgjm_result()` /
`bgjm_collect()` raise an informative error pointing at the job's logs (see
[Handling failures](#handling-failures)).

Convenient one-liner: pass `auto_remove = TRUE` to collect the result and drop
the job (and its working directory) from the registry in one step:

```{r}
#| eval: false
fit <- bgjm_result(job_id, auto_remove = TRUE)
```

# Examples

## SEM mediation with bootstrapped standard errors

Bootstrap procedures are ideal background jobs given their cost:

```{r}
#| eval: false
mediation_model <- "
  M ~ a * X
  Y ~ b * M + c * X
  indirect := a * b
  total := c + (a * b)
"

set.seed(123)
n <- 200
X <- rnorm(n); M <- 0.5 * X + rnorm(n); Y <- 0.3 * M + 0.2 * X + rnorm(n)
mediation_data <- data.frame(X, M, Y)

job_id <- bgjm_start_lavaan(
  model_syntax = mediation_model,
  data = mediation_data,
  lavaan_fun = "sem",
  job_name = "mediation",
  se = "bootstrap",
  bootstrap = 1000,
  seed = 42                # reproducible bootstrap
)

# ... keep working ...

fit <- bgjm_result(job_id, auto_remove = TRUE)
lavaan::parameterEstimates(fit, boot.ci.type = "bca.simple")
```

## Latent profile analysis with tidyLPA

```{r}
#| eval: false
library(tidyLPA)

job_id <- bgjm_start_tidylpa(
  data = iris[, 1:4],
  n_profiles = 3,
  models = 1,
  package = "mclust",
  job_name = "iris-lpa"
)

lpa <- bgjm_result(job_id, auto_remove = TRUE)
lpa
```

## Several models at once

Because jobs are non-blocking, launch competing models together and collect them
afterwards. The daemon pool caps how many actually run in parallel. With
`purrr`, starting and collecting are each one line:

```{r}
#| eval: false
library(purrr)

models <- list(
  "1-factor" = "g =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9",
  "2-factor" = "verbal =~ x1 + x2 + x3 + x4 + x5 + x6\n speed =~ x7 + x8 + x9",
  "3-factor" = "visual =~ x1 + x2 + x3\n textual =~ x4 + x5 + x6\n speed =~ x7 + x8 + x9"
)

# Start every model in the background; imap passes each syntax (.x) and its
# name (.y). Returns a named vector of job ids.
jobs <- imap_chr(models, \(syntax, name)
  bgjm_start_lavaan(syntax, HolzingerSwineford1939, "cfa", job_name = name))

# Collect them — each call waits until that job is ready.
fits <- map(jobs, bgjm_result)

tibble::tibble(
  model = names(fits),
  cfi   = map_dbl(fits, lavaan::fitMeasures, "cfi"),
  rmsea = map_dbl(fits, lavaan::fitMeasures, "rmsea"),
  aic   = map_dbl(fits, lavaan::fitMeasures, "aic")
)

walk(jobs, bgjm_remove, delete_dir = TRUE)
```

# Mplus jobs and file artifacts

Some tools — Mplus above all — produce *files* (`.inp`, `.dat`, `.out`). The job
system captures these for you. When a job runs, its directory looks like:

```
job_directory/
├── tmp/              # scratch (TMPDIR) for the job
├── artifacts/        # files written during the run land here
├── stdout.log        # captured standard output
└── stderr.log        # captured messages/warnings
```

## Running an Mplus input: `bgjm_start_mplus()`

`bgjm_start_mplus()` runs an existing Mplus input via
`MplusAutomation::runModels()`. Point it at a `.inp` file or a directory holding
the `.inp` (and its `.dat`); the inputs are staged into the job directory, Mplus
runs there, and the `.out` plus an `mplus_run.log` are captured as artifacts. The
collected `$result` is the parsed `MplusAutomation::readModels()` object.

```{r}
#| eval: false
job_id <- bgjm_start_mplus(
  target = "path/to/model_dir",   # holds model.inp (+ model.dat)
  job_name = "my-mplus-run"
)

out <- bgjm_collect(job_id)

list.files(out$artifacts_dir)     # e.g. "model.out", "mplus_run.log"
out$result                        # parsed model (readModels)

# Mplus streams its own console log (showOutput = TRUE, quiet = FALSE);
# on macOS/Linux it is captured to the job's stdout log:
head(bgjm_read_stdout(job_id))
```

To keep the output files permanently, pass a persistent `output_base` and remove
the job from the registry *without* deleting its directory:

```{r}
#| eval: false
job_id <- bgjm_start_mplus(
  target = "path/to/model_dir",
  output_base = file.path(getwd(), "mplus_results")
)
out <- bgjm_collect(job_id)
bgjm_remove(job_id, delete_dir = FALSE)   # keep files on disk
```

By default `output_base` is R's `tempdir()`, which may be cleared when R exits —
use a persistent `output_base` for anything you want to keep.

## LPA via the Mplus backend

`bgjm_start_tidylpa(package = "Mplus")` runs LPA through Mplus and captures the
same artifacts. Mplus requires alphanumeric/underscore variable names (no dots):

```{r}
#| eval: false
iris_mplus <- iris[, 1:4]
names(iris_mplus) <- c("Sepal_Length", "Sepal_Width",
                       "Petal_Length", "Petal_Width")

job_id <- bgjm_start_tidylpa(
  data = iris_mplus, n_profiles = 3, models = 1,
  package = "Mplus", keepfiles = TRUE,
  job_name = "iris-lpa-mplus"
)
result <- bgjm_collect(job_id)
list.files(result$artifacts_dir)
```

# Monitoring, logs, and control

```{r}
#| eval: false
bgjm_list()                       # registry snapshot (id, name, status, paths)

bgjm_status(job_id)               # non-blocking status

cat(bgjm_read_stdout(job_id), sep = "\n")   # captured stdout (incl. Mplus log)
cat(bgjm_read_stderr(job_id), sep = "\n")   # captured messages/warnings

bgjm_kill(job_id)                 # cancel a running job (mirai::stop_mirai)
bgjm_remove(job_id, delete_dir = TRUE)      # drop from registry (+ files)
```

## Job naming

A supplied `job_name` is used verbatim when unique, otherwise `-1`, `-2`, ... is
appended. With no name, a sensible default is generated (`"lavaan-cfa"`,
`"tidylpa-n3-mclust"`, `"mplus"`):

```{r}
#| eval: false
bgjm_start_lavaan(model_syntax, data, "cfa", job_name = "my-model")  # "my-model"
bgjm_start_lavaan(model_syntax, data, "cfa", job_name = "my-model")  # "my-model-1"
```

## Handling failures

A job that errors reports `status == "error"`, and `bgjm_collect()` raises an
error pointing at the logs:

```{r}
#| eval: false
result <- tryCatch(
  bgjm_collect(job_id),
  error = function(e) {
    message("Job failed: ", conditionMessage(e))
    cat(bgjm_read_stderr(job_id), sep = "\n")
    NULL
  }
)
bgjm_remove(job_id, delete_dir = TRUE)
```

# Best practices

- **Clean up in functions** with `on.exit()`:

  ```{r}
  #| eval: false
  analyze <- function(model, data) {
    job <- bgjm_start_lavaan(model, data, "cfa")
    on.exit(bgjm_remove(job, delete_dir = TRUE), add = TRUE)
    bgjm_collect(job)$result
  }
  ```

- **Use persistent `output_base`** for Mplus / anything whose files you want to
  keep; keep `tempdir()` for quick exploratory fits you collect immediately.
- **Use descriptive `job_name`s** so `bgjm_list()` is readable.
- **Size the pool** with `bgjm_daemons(n)` to match your cores and how many
  long jobs you expect to overlap.

# Notes and limits

- The registry lives in memory and is lost when R exits (mirai daemons are also
  shut down on package unload). "Background" means *another process this
  session*, not a job that survives a restart.
- `bgjm_collect()` blocks; use `bgjm_status()` / `bgjm_list()` for non-blocking
  checks.
- The Mplus console log is captured to stdout only where `runModels()` routes it
  through R (macOS/Linux). The authoritative record is always the `.out`
  artifact.
- For parallel *maps* with progress bars / early-stop, or timeouts, Shiny, and
  remote execution, prefer `mirai` directly.

# Session info

```{r}
#| eval: true
sessionInfo()
```
