---
title: "Formatting Mplus Coefficients with franzpak"
format:
  html:
    toc: true
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

This vignette demonstrates the Mplus coefficient tools in `franzpak`:

- `coef_wrapper()` -- pulls coefficients from an
  [`MplusAutomation`](https://cran.r-project.org/package=MplusAutomation)
  model and splits the Mplus parameter labels into dependent (`DV`) and
  independent (`IV`) variable columns.
- `coef_table_mplus()` -- arranges those coefficients into a publication-ready,
  predictor x outcome table (rendered with [`gt`](https://gt.rstudio.com) when
  available).

Both functions:

- auto-detect Bayesian vs. frequentist estimation and adapt the display
  (estimate + SE with *p*-value stars, or estimate + 95% interval with
  interval-exclusion stars);
- understand **two-level (multilevel) models**, splitting effects into
  **Within** and **Between** sections and reporting the means/intercepts of
  random slopes as the *between-person means of within-person effects*. When a
  model has random slopes, `coef_table_mplus()` returns a *pair* of tables --
  the fixed-effects table plus a dedicated **Random effects** table.

The examples use the classic Mplus User's Guide data files that ship with
`franzpak` in `inst/extdata/`.

```{r setup}
library(franzpak)
library(MplusAutomation)

# Locate a bundled .out file whether the package is installed or loaded from
# the source tree (e.g. via devtools::load_all()).
mplus_file <- function(name) {
  p <- system.file("extdata", name, package = "franzpak")
  if (nzchar(p)) return(p)
  cands <- c(file.path("inst", "extdata", name),
             file.path("..", "extdata", name),
             file.path("..", "..", "inst", "extdata", name))
  hit <- cands[file.exists(cands)]
  if (length(hit) == 0) stop("Could not locate ", name)
  hit[[1]]
}
```

# Part 1: A single-level model

`ex5.11` is a single-level structural equation model: two exogenous factors
(`F1`, `F2`) predict a mediator factor (`F3`), which in turn predicts an
outcome factor (`F4`).

```{r read-ex511}
m_sem <- readModels(mplus_file("ex5.11.out"), quiet = TRUE)
```

## `coef_wrapper()`: tidy coefficients

By default `coef_wrapper()` returns the unstandardized regression parameters
with the Mplus `DV<-IV` label split into separate columns.

```{r cw-basic}
coef_wrapper(m_sem)
```

Add confidence intervals with `addci = TRUE` (these come from the Mplus
CINTERVAL output):

```{r cw-ci}
coef_wrapper(m_sem, addci = TRUE)
```

Request a different coefficient scale with `type` -- e.g. fully standardized
(`STDYX`) estimates:

```{r cw-stdyx}
coef_wrapper(m_sem, type = "stdyx")
```

## `coef_table_mplus()`: a publication table

`coef_table_mplus()` reshapes the coefficients into a predictor x outcome
table. For a frequentist model the default display is estimate + standard error
with stars for *p* < 0.05.

```{r ct-est-se}
coef_table_mplus(m_sem)
```

Switch to an estimate + confidence-interval display with
`display_type = "est_ci"`; stars now mark intervals that exclude zero:

```{r ct-est-ci}
coef_table_mplus(m_sem, display_type = "est_ci")
```

Rename the cryptic Mplus factor labels with `label_replace`. Empty cells are
shown as dashes by default; set `na_replace = NULL` when you want true `NA`
values in the underlying data:

```{r ct-labels}
coef_table_mplus(
  m_sem,
  label_replace = c("F1" = "Predictor 1", "F2" = "Predictor 2",
                    "F3" = "Mediator",    "F4" = "Outcome")
)
```

For data manipulation instead of display, use `return_data = TRUE`:

```{r ct-raw}
coef_table_mplus(m_sem, return_data = TRUE, na_replace = NULL)
```

Control the order of the predictor rows with `predictor_order`. Listed
predictors come first, in the order given; any not listed keep their original
position after them. The names must match the *displayed* labels, so they apply
after `label_replace`:

```{r ct-order}
coef_table_mplus(
  m_sem,
  label_replace   = c("F1" = "Predictor 1", "F2" = "Predictor 2"),
  predictor_order = c("Predictor 2", "Predictor 1")
)
```

In two-level tables the order is applied within each section (Within / Between /
Random effects).

## Bayesian estimation is detected automatically

`ex5.11_BAYES` is the same model estimated with `ESTIMATOR = BAYES`. The tools
detect this, switch the default display to estimate + 95% **credibility**
interval, and emit a message reminding you that Mplus Bayesian *p*-values are
one-tailed.

```{r ct-bayes, message=TRUE}
m_sem_bayes <- readModels(mplus_file("ex5.11_BAYES.out"), quiet = TRUE)
coef_table_mplus(m_sem_bayes)
```

You can always override the auto-detection with the `bayes` and `display_type`
arguments.

# Part 2: A two-level model with a random slope

`ex9.2c` is a two-level regression (Mplus User's Guide example 9.2c). The model
is:

```default
%WITHIN%
  s | y ON x;        ! random slope of y on x (a within-person effect)
%BETWEEN%
  y s ON w x;        ! between-level regressions of y and of the slope s
  y WITH s;
```

The key feature is the **random slope** `s | y ON x`: the effect of `x` on `y`
*within* clusters is allowed to vary *between* clusters. Its intercept is the
average within-person slope -- a between-person summary of a within-person
effect.

```{r read-ex92c}
m_ml <- readModels(mplus_file("ex9.2c.out"), quiet = TRUE)
```

## `coef_wrapper()` annotates the multilevel structure

For two-level models `coef_wrapper()` adds several annotation columns. Here we
request `regression`, `expectation` and `variability` parameters so that
intercepts/means and the random-slope variance are all included:

```{r cw-ml}
coef_wrapper(m_ml, params = c("regression", "expectation", "variability"))
```

- **`level`** -- the Mplus section the parameter belongs to (`Within` /
  `Between`).
- **`random_slope`** -- `TRUE` for any parameter that pertains to the random
  slope `S` (its mean, its predictors, or its variance).
- **`within_avg_effect`** -- `TRUE` for the mean/intercept of a random slope,
  i.e. a between-person mean of a within-person effect.
- **`random_effect`** -- `TRUE` for the slope's *other* parameters: its
  cross-level predictors and its (residual) variance.
- **`variance`** -- `TRUE` for (residual) variance parameters.
- **`display_level`** -- the table section the row belongs to: `Within`,
  `Between`, or `Random effects`.

`coef_wrapper()` re-expresses each slope parameter in terms of the within-person
relationship the slope captures (`s | y ON x`):

- the **mean** of `S` (1.017) becomes the average within-person effect of `x` on
  `y`, so it is reported with `DV = "Y"`, `IV = "X"` and `display_level = "Within"`;
- the **cross-level regressions** `S ON W` and `S ON X` describe how the slope
  varies across clusters; they move to `display_level = "Random effects"` and are
  reported under the slope's outcome (`DV = "Y"`), keeping the between predictor;
- the slope's **(residual) variance** also moves to `Random effects`, labelled
  `IV = "Residual variance"`.

Variances that do not belong to a random slope (here the residual variances of
`Y`) are kept by `coef_wrapper()` but omitted from the table below.

## `coef_table_mplus()` returns a fixed-effects table and a Random effects table

A random slope is associated with a *specific within-person fixed effect* and
only secondarily with an outcome variable, so cramming all of its parameters
into the predictor x outcome grid (under the `Y` column) made the table hard to
read. Instead, when a model has random slopes `coef_table_mplus()` returns a
**named list** with two elements:

- `$fixed` -- the main predictor x outcome table, with **Within** and
  **Between** sections only;
- `$random` -- a **Random effects** table listing every random-slope parameter.

```{r ct-ml}
tbls <- coef_table_mplus(m_ml)
tbls$fixed
```

The fixed-effects table keeps two sections:

- **Within** -- within-person effects. The within-person effect of `X` on `Y`
  (the random-slope mean) appears as predictor `X` under outcome `Y`, marked with
  a superscript letter (the APA convention for a *specific note*; the dagger is
  avoided because APA reserves it for marginal significance).
- **Between** -- between-person effects on the real outcome `Y`. The
  *between-person* effect of `X` on `Y` (1.024) is a separate, fixed coefficient,
  kept clearly apart from its *within-person* counterpart (1.017). Cross-level
  interactions (the slope `S` regressed on a between predictor) stay here too,
  **relabelled to name the interaction**: `S ON W` becomes
  `W (between) x X (within)` (the between variable `W` moderating the within
  `X -> Y` slope).

The Random effects table collects the slope's parameters, grouped by slope and
ordered predictors -> mean/intercept -> variance. A slope's mean/intercept and
variance are relabelled by slope (`S Intercept`, `S Residual Variance`), while
its cross-level predictors keep their raw Mplus label (e.g. `S<-W`) so their
source stays unambiguous:

```{r re-ml}
tbls$random
```

Use `label_replace` to give the outcomes and predictors more descriptive names
in the fixed table (slope parameters follow their remapped predictor/outcome, so
renaming `X` and `Y` is enough; the relabelled cross-level interaction picks up
the new names too):

```{r ct-ml-labels}
coef_table_mplus(
  m_ml,
  label_replace = c("Y" = "Outcome Y", "X" = "Predictor X", "W" = "Cluster W")
)$fixed
```

When the package is rendering to plain tibbles (no `gt` installed), each table is
a tibble; the fixed table carries its section in a leading `level` column instead
of a row group, so the structure is never lost.

## Two-level Bayesian model

The Bayesian variant `ex9.2c_BAYES` defaults to a credibility-interval display.
The random-slope mean is again placed in the Within section, and the Random
effects table reports CrI limits.

```{r ct-ml-bayes, message=TRUE}
m_ml_bayes <- readModels(mplus_file("ex9.2c_BAYES.out"), quiet = TRUE)
ml_bayes <- coef_table_mplus(m_ml_bayes)
ml_bayes$fixed
ml_bayes$random
```

::: {.callout-note}
## Where the intervals come from

`coef_table_mplus()` (via an internal helper) first looks for interval columns
*in the main estimate table* -- the credibility limits Bayesian runs print
inline in `MODEL RESULTS`. These are reconstructed from the same parameter
columns that supply the estimates, so the labels always line up and the values
join cleanly even when variable names are long. Only if the estimate table
carries no intervals (typical of frequentist runs) does it fall back to
`MplusAutomation`'s `confint()`, which reads the dedicated `CINTERVAL` table. If
no intervals can be found at all (a frequentist run without `CINTERVAL`), the
table gracefully falls back to the estimate + SE display.

Preferring the estimate table also sidesteps an Mplus quirk: variable names are
truncated to 8 characters in the `CINTERVAL` section, so a long name like
`M_APREG_S` is printed there as `M_APREG_`. Joining intervals from that section
by label would silently drop the affected rows; reading from the (untruncated)
estimate table avoids the problem.
:::

# Part 3: A two-level model with *several* random slopes

`ex9.5` (Mplus User's Guide example 9.5, Bayesian variant) is a two-level
**path analysis** with two random slopes, so it exercises the same machinery as
Part 2 but with more than one slope in play:

```default
%WITHIN%
  s2 | y2 ON y1;     ! random slope of y2 on y1
  y2 ON x2;          ! fixed within-person effect
  s1 | y1 ON x2;     ! random slope of y1 on x2
  y1 ON x1;          ! fixed within-person effect
%BETWEEN%
  y1 y2 s1 s2 ON w;  ! between-level regressions of the outcomes and both slopes
```

The two slopes capture different within-person relationships -- `s2` is the
effect of `y1` on `y2`, and `s1` is the effect of `x2` on `y1` -- so each is
re-expressed under its own outcome (`Y2` and `Y1` respectively).

```{r read-ex95}
m_ml2 <- readModels(mplus_file("ex9.5_BAYES.out"), quiet = TRUE)
```

`coef_wrapper()` classifies the parameters of *both* slopes independently. The
latent slope names `S1`/`S2` never reach the output: each slope's mean, its
cross-level predictor `W`, and its residual variance are remapped to the
within-person relationship it represents.

```{r cw-ml2}
coef_wrapper(m_ml2, params = c("regression", "expectation", "variability"))
```

In the fixed-effects table, the means of `s2` (the within-person effect of `y1`
on `y2`) and `s1` (the within-person effect of `x2` on `y1`) appear in the
**Within** section marked with a superscript letter. Each slope's cross-level
predictor `W` becomes a relabelled interaction in **Between** (`W (between) x
Y1 (within)` under `Y2`, `W (between) x X2 (within)` under `Y1`):

```{r ct-ml2, message=TRUE}
tbls2 <- coef_table_mplus(m_ml2)
tbls2$fixed
```

The Random effects table lists both slopes' parameters, grouped by slope (`S1`
then `S2`) and ordered predictor -> mean -> variance. Means and variances are
relabelled by slope (`S1 Intercept`, `S1 Residual Variance`, ...); cross-level
predictors keep their Mplus label (`S1<-W`, `S2<-W`):

```{r re-ml2}
tbls2$random
```

# Part 4: Growth models and the `|` operator

The `|` operator serves multiple roles in Mplus. It declares **random slopes**
(Parts 2--3), but also **growth factors** and **latent variable interactions**
(`XWITH`). The package distinguishes these cases conservatively:

- growth-factor declarations such as `i s | y1-y4 AT ...` trigger the
  growth-specific table layout automatically; use `model_type = "growth"` to
  force it, or `model_type = "default"` for the ordinary predictor x outcome
  layout;
- ON-form random slopes are still handled by the two-level fixed/random table
  machinery in Parts 2--3;
- `XWITH` interactions are **not** random effects -- the interaction term is an
  ordinary predictor -- so they are detected and excluded from the random-effect
  handling (Part 5).

`ex6.12` is a single-level growth model with `TYPE = RANDOM`: two growth factors
(`i`, `s`) declared with `|`, a random slope `st` for a time-varying covariate,
and all three regressed on `x`:

```default
i s | y1-y4 AT a11-a14;   ! intercept & slope growth factors
st | y1 ON a21;           ! random slope of a time-varying covariate
st | y2 ON a22;
st | y3 ON a23;
st | y4 ON a24;
i s st ON x;              ! growth factors / slope regressed on x
```

```{r read-ex612}
m_growth <- readModels(mplus_file("ex6.12.out"), quiet = TRUE)
```

The slope detector recognises every `|` declaration, tagging the growth factors
`kind = "growth"` (no outcome/predictor decomposition) and the random slope
`st` `kind = "slope"`:

```{r rs-growth}
franzpak:::mplus_random_slopes(m_growth)
```

Because this model is **single-level** (no `%WITHIN%` / `%BETWEEN%` split), the
multilevel re-expression used in Parts 2--3 does not apply. `coef_wrapper()`
returns plain coefficients with each growth factor / slope keeping its own name
as the outcome -- the regression DVs are no longer corrupted by the random-slope
remapping:

```{r cw-growth}
coef_wrapper(m_growth)
```

`coef_table_mplus()` detects the growth-factor declaration and returns a
growth-specific table. The predictors of the factors are shown first, followed
by dedicated sections for factor means and variances:

```{r ct-growth}
coef_table_mplus(m_growth)
```

# Part 5: Latent variable interactions (`XWITH`)

`ex5.13` is a single-level SEM with a latent interaction. The `|` here defines an
interaction term `f1xf2` between two latent variables, which then predicts `f3`:

```default
f4 ON f3;
f3 ON f1 f2;
f1xf2 | f1 XWITH f2;   ! latent interaction term (NOT a random effect)
f3 ON f1xf2;
```

```{r read-ex513}
m_xwith <- readModels(mplus_file("ex5.13.out"), quiet = TRUE)
```

`XWITH` is detected by the `|` parser and **excluded** from the random effects,
so the helper returns an empty tibble:

```{r rs-xwith}
franzpak:::mplus_random_slopes(m_xwith)
```

The interaction term is kept as an ordinary predictor. `F1XF2` appears as a
predictor of `F3`, untouched by any random-effect remapping:

```{r cw-xwith}
coef_wrapper(m_xwith, params = "regression")
```

```{r ct-xwith}
coef_table_mplus(m_xwith)
```

Use `label_replace` to give the interaction a readable name:

```{r ct-xwith-labels}
coef_table_mplus(
  m_xwith,
  label_replace = c("F1XF2" = "F1 x F2")
)
```

# Argument quick reference

| Argument | `coef_wrapper()` | `coef_table_mplus()` |
|---|---|---|
| `params` | parameter types to extract (default `"regression"`) | default `c("regression", "expectation", "variability")` for two-level and growth tables, `"regression"` otherwise |
| `type` | coefficient scale: `"un"`, `"stdyx"`, `"stdy"`, `"std"` | passed through to `coef_wrapper()` |
| `bayes` | `NULL` auto-detect, or force `TRUE`/`FALSE` | same |
| `display_type` | -- | `NULL` auto-detect, `"est_se"`, or `"est_ci"` |
| `addci` | append `LowerCI`/`UpperCI` | (handled internally for `est_ci`) |
| `label_replace` | rename DV/IV labels | rename DV/IV labels |
| `na_replace` | -- | fill empty cells; default `"-"`, set `NULL` for true `NA` |
| `predictor_order` | -- | order predictor (row) labels |
| `model_type` | -- | `"auto"`, `"growth"`, or `"default"` |
| `return_data` | -- | return the underlying wide tibble(s) instead of `gt` |
| `sig_threshold`, `digits` | -- | star threshold / rounding |

# Session info

```{r session}
sessionInfo()
```
