---
title: "Formatting Mplus Coefficients with franzpak"
format:
  html:
    toc: true
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

This vignette demonstrates the Mplus coefficient tools in `franzpak`:

- `coef_wrapper()` -- pulls coefficients from an
  [`MplusAutomation`](https://cran.r-project.org/package=MplusAutomation)
  model and splits the Mplus parameter labels into dependent (`DV`) and
  independent (`IV`) variable columns.
- `coef_table_mplus()` -- arranges those coefficients into a publication-ready,
  predictor x outcome table (rendered with [`gt`](https://gt.rstudio.com) when
  available).

Both functions:

- auto-detect Bayesian vs. frequentist estimation and adapt the display
  (estimate + SE with *p*-value stars, or estimate + 95% interval with
  interval-exclusion stars);
- understand **two-level (multilevel) models**, splitting effects into
  **Within** and **Between** sections and reporting the means/intercepts of
  random slopes as the *between-person means of within-person effects*.

The examples use the classic Mplus User's Guide data files that ship with
`franzpak` in `inst/extdata/`.

```{r setup}
library(franzpak)
library(MplusAutomation)

# Locate a bundled .out file whether the package is installed or loaded from
# the source tree (e.g. via devtools::load_all()).
mplus_file <- function(name) {
  p <- system.file("extdata", name, package = "franzpak")
  if (nzchar(p)) return(p)
  cands <- c(file.path("inst", "extdata", name),
             file.path("..", "extdata", name),
             file.path("..", "..", "inst", "extdata", name))
  hit <- cands[file.exists(cands)]
  if (length(hit) == 0) stop("Could not locate ", name)
  hit[[1]]
}
```

# Part 1: A single-level model

`ex5.11` is a single-level structural equation model: two exogenous factors
(`F1`, `F2`) predict a mediator factor (`F3`), which in turn predicts an
outcome factor (`F4`).

```{r read-ex511}
m_sem <- readModels(mplus_file("ex5.11.out"), quiet = TRUE)
```

## `coef_wrapper()`: tidy coefficients

By default `coef_wrapper()` returns the unstandardized regression parameters
with the Mplus `DV<-IV` label split into separate columns.

```{r cw-basic}
coef_wrapper(m_sem)
```

Add confidence intervals with `addci = TRUE` (these come from the Mplus
CINTERVAL output):

```{r cw-ci}
coef_wrapper(m_sem, addci = TRUE)
```

Request a different coefficient scale with `type` -- e.g. fully standardized
(`STDYX`) estimates:

```{r cw-stdyx}
coef_wrapper(m_sem, type = "stdyx")
```

## `coef_table_mplus()`: a publication table

`coef_table_mplus()` reshapes the coefficients into a predictor x outcome
table. For a frequentist model the default display is estimate + standard error
with stars for *p* < 0.05.

```{r ct-est-se}
coef_table_mplus(m_sem)
```

Switch to an estimate + confidence-interval display with
`display_type = "est_ci"`; stars now mark intervals that exclude zero:

```{r ct-est-ci}
coef_table_mplus(m_sem, display_type = "est_ci")
```

Rename the cryptic Mplus factor labels with `label_replace`, and fill empty
cells with `na_replace`:

```{r ct-labels}
coef_table_mplus(
  m_sem,
  label_replace = c("F1" = "Predictor 1", "F2" = "Predictor 2",
                    "F3" = "Mediator",    "F4" = "Outcome"),
  na_replace    = "-"
)
```

## Bayesian estimation is detected automatically

`ex5.11_BAYES` is the same model estimated with `ESTIMATOR = BAYES`. The tools
detect this, switch the default display to estimate + 95% **credibility**
interval, and emit a message reminding you that Mplus Bayesian *p*-values are
one-tailed.

```{r ct-bayes, message=TRUE}
m_sem_bayes <- readModels(mplus_file("ex5.11_BAYES.out"), quiet = TRUE)
coef_table_mplus(m_sem_bayes)
```

You can always override the auto-detection with the `bayes` and `display_type`
arguments.

# Part 2: A two-level model with a random slope

`ex9.2c` is a two-level regression (Mplus User's Guide example 9.2c). The model
is:

```default
%WITHIN%
  s | y ON x;        ! random slope of y on x (a within-person effect)
%BETWEEN%
  y s ON w x;        ! between-level regressions of y and of the slope s
  y WITH s;
```

The key feature is the **random slope** `s | y ON x`: the effect of `x` on `y`
*within* clusters is allowed to vary *between* clusters. Its intercept is the
average within-person slope -- a between-person summary of a within-person
effect.

```{r read-ex92c}
m_ml <- readModels(mplus_file("ex9.2c.out"), quiet = TRUE)
```

## `coef_wrapper()` annotates the multilevel structure

For two-level models `coef_wrapper()` adds several annotation columns. Here we
request `regression`, `expectation` and `variability` parameters so that
intercepts/means and the random-slope variance are all included:

```{r cw-ml}
coef_wrapper(m_ml, params = c("regression", "expectation", "variability"))
```

- **`level`** -- the Mplus section the parameter belongs to (`Within` /
  `Between`).
- **`random_slope`** -- `TRUE` for any parameter that pertains to the random
  slope `S` (its mean, its predictors, or its variance).
- **`within_avg_effect`** -- `TRUE` for the mean/intercept of a random slope,
  i.e. a between-person mean of a within-person effect.
- **`random_effect`** -- `TRUE` for the slope's *other* parameters: its
  cross-level predictors and its (residual) variance.
- **`variance`** -- `TRUE` for (residual) variance parameters.
- **`display_level`** -- the table section the row belongs to: `Within`,
  `Between`, or `Random effects`.

`coef_wrapper()` re-expresses each slope parameter in terms of the within-person
relationship the slope captures (`s | y ON x`):

- the **mean** of `S` (1.017) becomes the average within-person effect of `x` on
  `y`, so it is reported with `DV = "Y"`, `IV = "X"` and `display_level = "Within"`;
- the **cross-level regressions** `S ON W` and `S ON X` describe how the slope
  varies across clusters; they move to `display_level = "Random effects"` and are
  reported under the slope's outcome (`DV = "Y"`), keeping the between predictor;
- the slope's **(residual) variance** also moves to `Random effects`, labelled
  `IV = "Residual variance"`.

Variances that do not belong to a random slope (here the residual variances of
`Y`) are kept by `coef_wrapper()` but omitted from the table below.

## `coef_table_mplus()` builds a three-section table

The table now has up to three sections:

- **Within** -- within-person effects. The within-person effect of `X` on `Y`
  (the random-slope mean) appears as predictor `X` under outcome `Y`, marked with
  a superscript letter (the APA convention for a *specific note*; the dagger is
  avoided because APA reserves it for marginal significance).
- **Between** -- between-person effects on the real outcome `Y`. Note that the
  *between-person* effect of `X` on `Y` (1.024) is a separate, fixed coefficient,
  kept clearly apart from its *within-person* counterpart (1.017) -- the classic
  within/between decomposition.
- **Random effects** -- the remaining parameters of the random slope `S`: its
  between-level predictors `W` and `X` (cross-level interactions) and its
  residual variance. These appear under the `Y` column instead of giving the
  latent slope `S` its own outcome column.

```{r ct-ml}
coef_table_mplus(m_ml)
```

Use `label_replace` to give the outcomes and predictors more descriptive names
(slope parameters follow their remapped predictor/outcome, so renaming `X` and
`Y` is enough):

```{r ct-ml-labels}
coef_table_mplus(
  m_ml,
  label_replace = c("Y" = "Outcome Y", "X" = "Predictor X", "W" = "Cluster W")
)
```

When the package is rendering to a plain tibble (no `gt` installed), the
section is carried in a leading `level` column instead of a row group, so the
section structure is never lost.

## Two-level Bayesian model

The Bayesian variant `ex9.2c_BAYES` defaults to a credibility-interval display.
The random-slope mean is again remapped to its within predictor and placed in
the Within section.

```{r ct-ml-bayes, message=TRUE}
m_ml_bayes <- readModels(mplus_file("ex9.2c_BAYES.out"), quiet = TRUE)
coef_table_mplus(m_ml_bayes)
```

::: {.callout-note}
## Intervals without a CINTERVAL table

`coef_table_mplus()` (via an internal helper) first asks
`MplusAutomation`'s `confint()` for intervals, which requires the Mplus
`CINTERVAL` output. If that table is unavailable -- common for Bayesian runs
where the credibility limits are printed inline in `MODEL RESULTS` -- the tools
fall back to reconstructing the 95% interval from those inline columns, so
`display_type = "est_ci"` keeps working. If no intervals can be found at all
(e.g. a frequentist run without `CINTERVAL`), the table gracefully falls back to
the estimate + SE display.
:::

# Argument quick reference

| Argument | `coef_wrapper()` | `coef_table_mplus()` |
|---|---|---|
| `params` | parameter types to extract (default `"regression"`) | default `c("regression", "expectation")` for two-level, `"regression"` otherwise |
| `type` | coefficient scale: `"un"`, `"stdyx"`, `"stdy"`, `"std"` | passed through to `coef_wrapper()` |
| `bayes` | `NULL` auto-detect, or force `TRUE`/`FALSE` | same |
| `display_type` | -- | `NULL` auto-detect, `"est_se"`, or `"est_ci"` |
| `addci` | append `LowerCI`/`UpperCI` | (handled internally for `est_ci`) |
| `label_replace` | rename DV/IV labels | rename DV/IV labels |
| `na_replace` | -- | fill empty cells |
| `sig_threshold`, `digits` | -- | star threshold / rounding |

# Session info

```{r session}
sessionInfo()
```
